Claret: Using Data Types for Highly Concurrent Distributed Transactions

### 2015-02-09 rmd/paper.Rmd
rmd: update styling & spec

\begin{table}
\centering
\begin{tabular}{lll}
\textbf{method:} & \textbf{commutes with:} & \textbf{when:} \\
\hline
\texttt{add(x) -> void} & \texttt{add(y)} & $\forall x, y$ \\
\texttt{remove(x) -> void} & \texttt{remove(y)} & $\forall x, y$ \\
    & \texttt{add(y)} & $x \ne y$ \\
\texttt{size() -> int} & \texttt{add(x)} & $x \in Set$ \\
    & \texttt{remove(x)} & $x \notin Set$ \\
\texttt{getall() -> list} & \texttt{add(x)} & $x \in Set$ \\
    & \texttt{remove(x)} & $x \notin Set$ \\
    & \texttt{size()} & $always$ \\
    \hline
\end{tabular}
\caption{\label{spec} Commutativity Specification for Set.}
\end{table}

Though *commutativity* is often discussed in terms of an operation commuting with all other operations, it is actually more nuanced. If a pair of operations commute, then executing them in either order will produce the same result. Using the definitions from \cite{Kulkarni:PLDI11}, whether or not a pair of method invocations commute is a property of the data type and is a function of the methods, their arguments, their return values, and the *abstract* state of their target. We call the full set of commutativity rules for a data type its *commutativity specification.* An example specification for a *Set* is shown in \autoref{tab:spec}. It is important to note that interface choice affects commutativity: if `add(x)` returned a boolean that expressed if the item was added or not, then `add(x)` would only commute with `add(y)` if $x \ne y$. In this case, the difference doesn't drastically impact commutativity, but it would affect the cost of checking commutativity dynamically.

For a given data type interface, there is one specification that expresses the maximum concurrency, like the one in \autoref{tab:spec}, but there may be many that express varying degrees of commutativity. The space of possible specifications forms a *lattice* where $\top$ specifies the maximum possible commutativity, and $\bot$ represents the case where no operations are allowed to commute. As Kulkarni \cite{Kulkarni:PLDI11} explored, there is a complicated tradeoff between exposing more concurrency and efficiently tracking commutativity. Data structure designers and users navigate these choices to optimize their programs.

@@

### Record splitting for parallelism

Doppel \cite{Narula:OSDI14} observed that several contentious operations, such as incrementing counters or keeping track of the maximum bid, actually commute because the application doesn't need the output of each update operation. This allows them to *split* hot records onto multiple cores and execute commutative operations in parallel, *reconciling* them back to a single value before executing reads or non-commuting operations. This observation can be generalized via ADTs: operations can operate on copies of a record in parallel provided they commute with each other and provided they can be combined at the end of a phase before beginning to execute a new phase with a different set of operations that didn't commute with the first set.

### Combining

Another way to reduce contention on a shared data structure is to synchronize hierarchically: first with a couple immediate neighbors, then possibly with more clusters of neighbors, and finally with the data structure itself. This is known as combining \cite{flatCombining,yew:combining-trees,funnels} in the multi-processing literature, and could be applied to contended records in our model just as well as to shared data structures where it originated.

>>>>

\begin{table}
\centering
\begin{tabular}{lll}
\textbf{method:} & \textbf{commutes with:} & \textbf{when:} \\
\hline
\texttt{add(x): void} & \texttt{add(y)} & $\forall x, y$ \\
\texttt{remove(x): void} & \texttt{remove(y)} & $\forall x, y$ \\
    & \texttt{add(y)} & $x \ne y$ \\
\texttt{size(): int} & \texttt{add(x)} & $x \in Set$ \\
    & \texttt{remove(x)} & $x \notin Set$ \\
\texttt{contains(x): bool} & \texttt{add(y)} & $x \ne y \lor y \in Set$ \\
    & \texttt{remove(y)} & $x \ne y \lor y \notin Set$ \\
    & \texttt{size()} & $\forall x$ \\
    \hline
\end{tabular}
\caption{\label{spec} Commutativity Specification for Set.}
\end{table}

Though *commutativity* is often discussed in terms of an operation commuting with all other operations, it is actually more nuanced. If a pair of operations commute, then executing them in either order will produce the same result. Using the definitions from \cite{Kulkarni:PLDI11}, whether or not a pair of method invocations commute is a property of the data type and is a function of the methods, their arguments, their return values, and the *abstract* state of their target. We call the full set of commutativity rules for a data type its *commutativity specification.* An example specification for a *Set* is shown in \autoref{spec}. It is important to note that interface choice affects commutativity: if `add(x)` returned a boolean that expressed if the item was added or not, then `add(x)` would only commute with `add(y)` if $x \ne y$. In this case, the difference doesn't drastically impact commutativity, but it would affect the cost of checking commutativity dynamically.

For a given data type interface, there is one specification that expresses the maximum concurrency, like the one in \autoref{spec}, but there may be many that express varying degrees of commutativity. The space of possible specifications forms a *lattice* where $\top$ specifies the maximum possible commutativity, and $\bot$ represents the case where no operations are allowed to commute. As Kulkarni \cite{Kulkarni:PLDI11} explored, there is a complicated tradeoff between exposing more concurrency and efficiently tracking commutativity. Data structure designers and users navigate these choices to optimize their programs.

@@

**Record splitting for parallelism.**
Doppel \cite{Narula:OSDI14} observed that several contentious operations, such as incrementing counters or keeping track of the maximum bid, actually commute because the application doesn't need the output of each update operation. This allows them to *split* hot records onto multiple cores and execute commutative operations in parallel, *reconciling* them back to a single value before executing reads or non-commuting operations. This observation can be generalized via ADTs: operations can operate on copies of a record in parallel provided they commute with each other and provided they can be combined at the end of a phase before beginning to execute a new phase with a different set of operations that didn't commute with the first set.

**Combining.**
Another way to reduce contention on a shared data structure is to synchronize hierarchically: first with a couple immediate neighbors, then possibly with more clusters of neighbors, and finally with the data structure itself. This is known as combining \cite{flatCombining,yew:combining-trees,funnels} in the multi-processing literature, and could be applied to contended records in our model just as well as to shared data structures where it originated.
